---
title: "Linear Modeling of Average Temperature Using Climate Change Indicators"
author: "Julia Ngo, Giuseppe Salvatori, and Layan Alhafi"
date: "`r Sys.Date()`"
output: pdf_document
---

# Introduction

This project analyzes the *[Global Climate Indicators: A Comprehensive Dataset (2000–2024)](https://www.kaggle.com/datasets/bhadramohit/climate-change-dataset)* sourced from Kaggle, which compiles environmental and climate-related data across multiple countries over time. The dataset includes key variables such as year, country, average temperature, CO2 emissions, sea level rise, rainfall, population, renewable energy usage, extreme weather events, and forest area, with approximately 1,000 observations and 10 variables in total.

This dataset is particularly relevant for exploring patterns and relationships within global climate change. Its structure makes it suitable for examining how average temperature is associated with other environmental and socioeconomic factors, as well as for comparing trends across countries and time periods. By analyzing these variables, the project aims to better understand how human activity and environmental changes interact to influence global temperature trends.

# Methods

Exploratory data analysis will be conducted to identify variables associated with average temperature. This will include generating scatterplots for each predictor variable against average temperature and calculating correlation coefficients to assess the strength and direction of relationships.

Multiple linear models will then be developed and compared. A multiple linear regression model will serve as the initial approach. If this model is inadequate, alternative methods will be considered, including physics-based models such as General Circulation Models (GCMs) and AI-driven models like Pangu-Weather and GenCast. In addition, single-variable regression models will be evaluated to determine whether individual predictors better explain variation in average temperature than a multivariable approach.

Model performance will be assessed using diagnostic plots, including residuals versus fitted values and Q-Q plots, to evaluate assumptions of linearity, homoscedasticity, and normality. If these assumptions are violated, alternative models will be prioritized. Models will be compared using R^2 values, with the highest value indicating the best fit.

```{r}
library(readr)
climate <- read.csv("climate_change_dataset.csv", fileEncoding = "UTF-8-BOM",
                    stringsAsFactors = FALSE)
```

## Exploratory Analysis

```{r}
names(climate) <- iconv(names(climate), from = "", to = "UTF-8", sub = "")
```

```{r message= FALSE}
library(dplyr)

tidy_climate <- climate %>%
  rename(
    year = Year,
    country = Country,
    avg_temp = Avg.Temperature...C.,
    co2_emissions = CO2.Emissions..Tons.Capita.,
    sea_level_rise = Sea.Level.Rise..mm.,
    rainfall = Rainfall..mm.,
    population = Population,
    renewable_energy = Renewable.Energy....,
    extreme_weather = Extreme.Weather.Events,
    forest_area = Forest.Area....
  )
```

```{r}
library(ggplot2)
library(tidyr)

long_climate <- tidy_climate %>%
  pivot_longer(
    cols = co2_emissions:forest_area,
    names_to = "climate_indicator",
    values_to = "value"
  )
```

```{r, fig.width=10, fig.height=10}
long_climate %>% ggplot() +
  geom_point(mapping=aes(x=value, y=avg_temp)) +
  facet_wrap(~climate_indicator, scales="free_x") +
  geom_smooth(mapping=aes(x=value, y=avg_temp), method="lm") +
  labs(title="Average Temp. vs. Climate Change Indicators (2000-2023)",
       x="Climate Indicators", y="Average Temp.") +
  scale_x_log10() +
  scale_y_log10()
```

```{r, fig.width=10, fig.height=10}
long_climate %>%
  filter(year >= 2018 & year <= 2023) %>%
  ggplot() +
  geom_point(mapping=aes(x=value, y=avg_temp)) +
  facet_wrap(~climate_indicator, scales="free_x") +
  geom_smooth(mapping=aes(x=value, y=avg_temp), method="lm") +
  labs(title="Average Temp. vs. Climate Change Indicators (2018-2023)",
       x="Climate Indicators", y="Average Temp.") +
  scale_x_log10() +
  scale_y_log10()
```

```{r, fig.width=10, fig.height=10}
long_climate %>%
  filter(country == "USA") %>%
  ggplot() +
  geom_point(aes(x = value, y = avg_temp)) +
  facet_wrap(~climate_indicator, scales = "free_x") +
  geom_smooth(aes(x = value, y = avg_temp), method = "lm") +
  labs(
    title = "Average Temp. vs. Climate Indicators (USA, 2000-2023)",
    x = "Climate Indicators",
    y = "Average Temp."
  ) +
  scale_x_log10() +
  scale_y_log10()
```

## Model

```{r}
library(broom)
US_climate <- tidy_climate %>%
  filter(country == "USA")
```

```{r}
US_climate_all <- lm(log(avg_temp) ~ log(co2_emissions) +
                       log(sea_level_rise) +
                       log(rainfall) +
                       log(population) +
                       log(renewable_energy) +
                       extreme_weather +
                       log(forest_area),
                     data = US_climate)
glance(US_climate_all)[1]
```

```{r}
US_climate_significant <- lm(log(avg_temp) ~ log(rainfall) +
                       log(population) +
                       extreme_weather +
                       log(forest_area),
                     data = US_climate)
glance(US_climate_significant)[1]
```

```{r}
US_temp_rainfall <- lm(log(avg_temp) ~ log(rainfall),data = US_climate)
glance(US_temp_rainfall)[1]
```

```{r}
US_temp_population <- lm(log(avg_temp) ~ log(population),data = US_climate)
glance(US_temp_population)[1]
```

```{r}
US_temp_extreme <- lm(log(avg_temp) ~ extreme_weather,data = US_climate)
glance(US_temp_extreme)[1]
```

```{r}
US_temp_forest <- lm(log(avg_temp) ~ log(forest_area),data = US_climate)
glance(US_temp_forest)[1]
```

## Linearity Assumptions

```{r}
library(modelr)

US_climate_t <- US_climate %>%
  add_predictions(US_climate_all) %>%
  add_residuals(US_climate_all)
```

```{r}
US_climate_t %>% 
  ggplot() +
  geom_point(mapping = aes(x = pred, y = log(avg_temp))) +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "US Climate Model: Observed vs Predicted Average Temperature (Log Scale)",
    x = "Predicted",
    y = "Observed"
  )
```

```{r}
US_climate_t %>% 
  ggplot() +
  geom_point(mapping = aes(x = pred, y = resid)) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Residuals vs Predicted (US Climate Model)",
    x = "Predicted",
    y = "Residuals"
    )
```

```{r}
US_climate_t %>%
  ggplot() +
  geom_qq(aes(sample = resid)) +
  geom_qq_line(aes(sample = resid)) +
  labs(title = "Q-Q Plot of Residuals (US Climate Model)")
```

```{r}
US_climate_significant_t <- US_climate %>%
  add_predictions(US_climate_significant) %>%
  add_residuals(US_climate_significant)
```

```{r}
US_climate_significant_t %>% 
  ggplot() +
  geom_point(mapping = aes(x = pred, y = log(avg_temp))) +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "Significant Variables Model: Predicted vs Observed Average Temp. (Log Scale)", 
    x = "Predicted", 
    y = "Observed"
  )
```

```{r}
US_climate_significant_t %>% 
  ggplot() +
  geom_point(mapping = aes(x = pred, y = resid)) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Residuals vs Predicted (Significant Variables Model)",
    x = "Predicted",
    y = "Residuals"
    )
```

```{r}
US_climate_significant_t %>%
  ggplot() +
  geom_qq(aes(sample = resid)) +
  geom_qq_line(aes(sample = resid)) +
  labs(title = "Q-Q Plot of Residuals (Significant Variables Model)")
```

```{r}
US_rainfall_t <- US_climate %>%
  add_predictions(US_temp_rainfall) %>%
  add_residuals(US_temp_rainfall)

US_rainfall_t %>%
ggplot() +
  geom_point(aes(x = log(rainfall), y = log(avg_temp))) +
  geom_abline(
    slope = US_temp_rainfall$coefficients[2],
    intercept = US_temp_rainfall$coefficients[1]
  ) +
  labs(
    title = "Average Temperature vs Rainfall (Log Scale)",
    x = "Rainfall",
    y = "Average Temperature"
  )

US_rainfall_t %>%
  ggplot() +
  geom_point(aes(x = pred, y = resid)) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Residuals vs Predicted (Rainfall Model)",
    x = "Predicted",
    y = "Residuals"
  )

US_rainfall_t %>%
  ggplot() +
  geom_qq(aes(sample = resid)) +
  geom_qq_line(aes(sample = resid)) +
  labs(title = "Q-Q Plot of Residuals (Rainfall Model)")
```

```{r}
US_extreme_t <- US_climate %>%
  add_predictions(US_temp_extreme) %>%
  add_residuals(US_temp_extreme)

US_extreme_t %>%
ggplot() +
  geom_point(aes(x = extreme_weather, y = log(avg_temp))) +
  geom_abline(
    slope = US_temp_extreme$coefficients[2],
    intercept = US_temp_extreme$coefficients[1]
  ) +
  labs(
    title = "Average Temperature (Log Scale) vs Extreme Weather Events",
    x = "Extreme Weather Events",
    y = "Average Temperature"
  )

US_extreme_t %>%
  ggplot() +
  geom_point(aes(x = pred, y = resid)) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Residuals vs Predicted (Extreme Weather Model)",
    x = "Predicted",
    y = "Residuals"
  )

US_extreme_t %>% 
 ggplot() +
  geom_qq(aes(sample = resid)) +
  geom_qq_line(aes(sample = resid)) +
  labs(title = "Q-Q Plot of Residuals (Extreme Weather Model)")
```

# Results

Across all scatterplots depicting the relationship between average temperature and a certain climate change indicator (2000-2023), there was no apparent relationship between these variables. In each case, the points were widely scattered with no clear direction or pattern -- indicating a weak or nonexistent association. When filtering the data to only include points from the past five years (2018-2023), a few variables were shown to have a relationship with average temperature. Population, the number of extreme weather events, and forest area exhibited weak positive relationships, while CO2 emissions, rainfall, and renewable energy showed weak negative relationships. Finally, when focusing on data only from the United States (2000-2023) -- several stronger relationships emerged. Extreme weather events had a fairly strong positive relationship with average temperature, as well as renewable energy. Rainfall, on the other hand, demonstrated a strong negative relationship, while forest area maintained a relatively weak association.

Using the above information/filtered data, multiple linear models were made; one of each significant variable alone, another using all of the significant variables and finally one containing all variables tested, regardless of significance. The linear model consisting of all variables as expected had the highest r^2 value of 0.1902673, which is considered weak/shows a weak correlation between the response variable and the explanatory variables. The linear model consisting of significant variables (rainfall, population, extreme weather events and forest area) had a very close r^2 value of 0.1865008. The variable with the strongest/highest r^2 was rainfall with 0.07761612, while the weakest variable was population with 0.003274811.

Then, linear model assumption plots were used to evaluate whether the linear models satisfied the assumptions required for a reliable linear model. For the full model and the reduced model containing only significant variables, the linear model assumption plots are nearly identical. For both models, the observed versus predicted plots showed a general positive relationship, but the points were widely scattered around the reference line, indicating that the linearity assumption is not satisfied. For both models, the residuals versus predicted plots showed residuals randomly scattered around zero with no clear curve or funnel shape. However, the residuals displayed a fairly wide range of variation, with the points ranging from approximately -1.0 to over 0.5. So, this suggests that the constant variability assumption is only partially satisfied. The Q-Q plots for both models showed that the residuals generally followed the reference line in the middle region with some deviation at the tails, indicating that the residuals are nearly normal. Similar evaluations were done on the two strongest single-variable models, rainfall and extreme weather. Both single-variable models have similar results to the full and significant models, although these single-variable models showed even weaker linear relationships. Overall, the assumption plots suggest that while the models capture some general trends in the data, a linear model does not fully explain the relationship between climate indicators and average temperature.


# Conclusion
Write your conclusion here.